WHITE WINE QUALITY by ABIR PATTNAIK
========================================================

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.
library(ggplot2)
library(gridExtra)
library(ggthemes)
library(Hmisc)
library(knitr)
library(corrplot)
library(quantreg)


# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.


```

```{r echo=FALSE, Load_the_Data}
# Load the Data
White_wine<-read.csv("C:/Users/DRDO HQ/Desktop/UDACITY NANODEGREE/P4 EDA/wineQualityWhites.csv") 

```

# INTRODUCTION

This dataset has been collected from:
  P. Cortez, A. Cerdeira, F. Almeida, T. Matos and J. Reis. 
  Modeling wine preferences by data mining from physicochemical properties.
  In Decision Support Systems, Elsevier, 47(4):547-553. ISSN: 0167-9236.
The data set is about the White wine and its quality.Here __Quality__ is the 
target variable and it has various factors that affect the White Wine quality.
This [site](http://www.vinhoverde.pt/en/) gives details of the palce where it 
was collected.

## About White Wine
White wine is a wine whose colour can be straw-yellow, yellow-green, or 
yellow-gold. It is produced by the alcoholic fermentation of the non-coloured
pulp of grapes,which may have a skin of any colour. White wine has existed 
for at least 2500 years.[Source](https://en.wikipedia.org/wiki/White_wine)

# Univariate Plots Section

For Univariate analysis, help of box plots and histograms were taken.Although 
histograms demonstrate distributions better but box plots help in identifying 
the outliers.

Both of these plots demonstrate their unique features hence can help in making 
one variate plots a little easier.Here is a [link](http://www.brighthubpm.com
/six-sigma/58254-box-plots-vs-histograms-in-project-management/)

```{r echo=FALSE, Univariate_Plots}
# A function was created to plot histogram and box plot
UnivariatePlots<-function(variable_factor){
  
  hist_plot<-ggplot(data=White_wine,
                    aes(variable_factor))+
      geom_histogram(color='black',
                     fill='#099DD9')+
      labs(x=deparse(substitute(variable_factor)),y='Total Count',
           title='Univariate plots of variable',
           subtitle='Histogram plot of variable')
 
  hist_plot_final<-hist_plot+theme_economist()+
                    scale_color_economist()
      
  box_plot<-ggplot(data=White_wine,aes(x=variable_factor,
                                       y=variable_factor))+
        geom_boxplot(outlier.color = 'green',
                     color='#099DD9',
                     fill='#FFA533')+
        labs(x=deparse(substitute(variable_factor)),y='Total Count',
             subtitle='Boxplot plot of variable')
  box_plot_final<-box_plot+theme_economist()+scale_color_economist()
  
  grid.arrange(hist_plot_final
               ,box_plot_final,
               nrow=2)
}
View(head(White_wine))
str(White_wine)
attach(White_wine)
```


```{r Fixed Acidity,warning=FALSE,message=FALSE}
UnivariatePlots(fixed.acidity)
describe(fixed.acidity)
```

# Fixed Acidity

1. The fixed acidity is coming at a mean of 6.85 g/dm^3^
2. Most of the wine has fixed acidity is between 6.3 g/dm^3^.
3. The hishest fixed acidity is 14.2 g/dm^3^ and is an outlier. 

```{r Volatile Acidity,warning=FALSE,message=FALSE}
UnivariatePlots(volatile.acidity)
describe(volatile.acidity)
```

# Volatile Acidity

The volatile acidity is the amount of acetic acid in wine.**High** values of 
lead to unpleasant smell.The histogram drawn is a left skewed graph with 
0.26 g/dm^3^ being the most quantity used in wine.

The last five outliers 0.91-1.10 g/dm^3^ lead in unpleasantness and needs to be
investigated further.The mean(0.278) is slightly higher than the 
median(0.26) due to the graph being left skewed.

```{r Citric Acid,warning=FALSE,message=FALSE}
UnivariatePlots(citric.acid)
describe(citric.acid)
```

# Citric Acid

Citric Acid is found in small quantities but it adds 'freshness' and flavor to 
wines.Thus most of them are in the range of 0.27 g/dm^3^ to 0.39 g/dm^3^.It is 
possible having 0 g/dm^3^ while it can also go high as 1.66 g/dm^3^.

```{r Residual Sugar,warning=FALSE,message=FALSE}
UnivariatePlots(residual.sugar)
describe(residual.sugar)
```

# Residual Sugar

Its the amount of sugar remaining after fermentation stops, it's rare to find 
wines with less than 1 gram/liter and wines with greater than 45 grams/liter 
are considered sweet.

Only 5% have 1.1g/dm^3^. while 95% is within 15.7 g/dm^3^.The outliers such as 
65.80 g/dm^3^ would be too sweet and its quality should be known.

```{r Chlorides,warning=FALSE,message=FALSE}
UnivariatePlots(chlorides)
describe(chlorides)
```

# Chlorides

Chlorides is the amount of salt in wine.Both the plots give interesting insight 
on the chlorides column. 95% of the values of chloride are within 0.067 g/dm^3^.
Thus high values _may_ not be suitable.If the outliers are removed the graph 
would then be a normal distribution curve.

```{r Free SUlphur Dioxide,warning=FALSE,message=FALSE}
UnivariatePlots(free.sulfur.dioxide)
describe(free.sulfur.dioxide)
```

# Free Sulphur Dioxide

The free form of SO~2~ exists in equilibrium between molecular SO~2~ (as a 
dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation
of wine.Most of them are in range of 23 mg/dm^3^ to 46 mg/dm^3^ but it goes as 
high as 289 g/dm^3^ and needs to investigated further.

```{r Total Sulphur Dioxide,warning=FALSE,message=FALSE }
UnivariatePlots(total.sulfur.dioxide)
describe(total.sulfur.dioxide)
```

#Total Sulphur Dioxide

It is the amount of free and bound forms of S0~2~; in low concentrations, SO~2~ 
is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO~2~
becomes evident in the nose and taste of wine.

Information needs to be clearer as most of the wines are in range of 
108 mg/dm^3^ to 167 mg/dm^3^.

```{r Density,warning=FALSE,message=FALSE}
UnivariatePlots(density)
describe(density)
```

# Density

As the mean and median are quite close together the outliers 1.03898 g/cm^3^ 
doesn't change the graph of histogram and most of them are in range of 
0.9917-0.9961.

The density of water is close to that of water depending on the percent alcohol 
and sugar content.

```{r pH,warning=FALSE,message=FALSE}
UnivariatePlots(pH)
describe(pH)
```

# pH

This is a fairly normal distribution curve. 95% pf values are below 3.46.pH 
describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 
(very basic).Most of them are between 3-4.

```{r Sulphate,warning=FALSE,message=FALSE}
UnivariatePlots(sulphates)
describe(sulphates)
```

# Sulphates

Although they aren't any extreme outliers most of sulphates are between 0.41 and
0.55 g/dm^3^.Sulphate is a wine additive which can contribute to sulfur dioxide 
gas (S0~2~) levels, wich acts as an antimicrobial and antioxidant.

```{r Alcohol,warning=FALSE,message=FALSE}
UnivariatePlots(alcohol)
describe(alcohol)
```

# Alcohol

The Histogram for alcohol is nothing like the plots of the previous variables as 
most of the values are spreaded over values hence it is inconclusive on whether 
it could be considered as a factor for quality or not.
It gives the the percent alcohol content of the wine.

```{r Quality,warning=FALSE,message=FALSE}
UnivariatePlots(quality)
describe(quality)
```

# Quality

Quality here is the target variable.We need to know what factors make the 
quality of wine 8 or higher.FOr these,univariate plots were created for each 
factor.The plots of Quality is a normal distribution curve with mean at 5.878.
We need to find relations with them and make an analysis further. 

# Categorising Quality variable

```{r category,message=FALSE,warning=FALSE}
# Creating variables to categorise the quality of wine into 4 groups namely,
# Bad,Average,Good,Excellent
category=cut(quality,4,labels=c('Bad','Average','Good','Excellent'))
summary(category)
barchart(table(category), freq=TRUE, 
         xlab = "Frequencies", 
         ylab = "Labels")
```

Average has the highest quantity of wines as compared to Bad and Excellent Wines.
Hence, we need to see what all qualities does these labels for bad and excellent 
wines are made.

# Univariate Analysis


### What is the structure of your dataset?
The structure of dataset is the white wine dataset in which there are 11 
variables that can help in determining the quality variable that is the target 
variable.There are in total of 4989 observations and has has been collected from
a winery in Western Portugal.The Quality can be converted into categorical 
variable while others are all continuos variables.

### What are the main feature(s) of interest in your dataset?

1.Our main objective is to determine which factors can __cause__ the quality of 
wine.
2.The alcohol column is inconclusive and needs to be researched upon that.
3.Another feature was the outliers in the data set and it needs to be researched 
upon with the Quality.
Quality here is a target variable and is of interest.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Features like which would be better for the test of wine such as volatile acidity
,citric acid, the amount of residual sugar may help in the investigation.

### Did you create any new variables from existing variables in the dataset?

Since the quality variable contains only integers I can convert them to 
categories like 'bad','average','good','great'.


# Bivariate Plots Section


```{r Bivariate_Plots,warning=FALSE,message=FALSE,fig.width=7.5,fig.height=7.5}
corrplot(cor(White_wine),method="number"
         ,type = "lower"
         ,order = "FPC"
         ,number.cex=1)
```

# Bivariate Analysis

Corrplot packages gives correlation between the variables and it does give 
interesting insights on how to variables give us features of interest.However 
it should be noted that __CORRELATION IS NOT CAUSATION__. This means that even 
though  variables may be correlated but it isn't necessary that it might be the 
cause.More understanding can be found by clicking this
[link](https://goo.gl/Trxp3W)

But there is a need of exploration on variables where they are correlated.
Therefore plots were created of the following graphs on the basis of correlation.
The **Rule of Thumb** says if two variables are correlated with >0.3 and <-0.3 
then it is meaningful.
A correlation of |0.5| is moderate.
A correlation of |0.7| is large.
Keeping that in mind following variables were investigated.
1.Alcohol v/s Chloride
2.Alcohol v/s Total Sulphuric Dioxide
3.Alcohol v/s residual sugar
4.Alcohol v/s Density
5.Density v/s Residual Sugar
6.Total Sulphuric Dioxide v/s Free Sulphuric Dioxide
7.Density v/s Total Sulphuric DIoxide
8.Quality v/s Alcohol

```{r Alcohol vs Chloride, warning=FALSE,message=FALSE}
ggplot(data=White_wine,aes(x=alcohol,y=chlorides))+
  geom_point(alpha=1/50)+
  geom_smooth()+
  coord_cartesian(ylim=c(0,0.1))+
  theme_economist()+
  xlab('Alcohol Percentage') +
  ylab('Chlorides') +
  labs(title = 'Alcohol Vs Chloride')
```

The alcohol content with 9% has chloride of approx 0.56 g/dm^3^.Then the graph 
has a fairly linear decrease as the alchol is increased.
```{r Alcohol vs TSD,warning=FALSE,message=FALSE}
ggplot(data=White_wine,aes(x=alcohol,y=total.sulfur.dioxide))+
  geom_point(alpha=1/20)+
  geom_quantile(color = 'navyblue')+
  xlab('Alcohol Percentage') +
  ylab('Total Sulphur Dioxide') +
  labs(title = 'Alcohol Percentage Vs Total Sulphur Dioxide')
```

The Quantreg package plots quantiles onto the graph as a layer onto the graph 
and displayed that 9% alcohol has a wide range of quantile and most of the wines
is present in this range only.

```{r Alcohol vs residual sugar,warning=FALSE,message=FALSE}
ggplot(data=White_wine,aes(x=alcohol,y=residual.sugar))+
  geom_point(alpha=1/20)+
  geom_smooth()+
  xlab('Alcohol Percentage') +
  ylab('Residual Sugar') +
  labs(title = 'Alcohol Percentage Vs Residual Sugar')
```

The residual sugar displays steep decline till it reaches 10% alcohol then there
is a stagnant decrease.

```{r Alcohol vs Density,warning=FALSE,message=FALSE}
ggplot(data=White_wine,aes(x=alcohol,y=density))+
  geom_point(alpha=1/50)+
  geom_smooth()+
  xlab('Alcohol Percentage') +
  ylab('Density') +
  labs(title = 'Alcohol Percentage Vs Density')
```

Drawing a line over a graph provides a fair relationship between  variables.In 
this there is a downward a linear relationship between density and alcohol.
Denser points are created between 9.6-9.8% alcohol.

```{r Density vs Residual Sugar,warning=FALSE,message=FALSE}
ggplot(data=White_wine,aes(x=residual.sugar,y=density))+
  geom_point(alpha=1/50)+
  geom_smooth()+
  coord_cartesian(xlim = c(0,20))+
  theme_economist()+
  xlab('Residual Sugar') +
  ylab('Density') +
  labs(title = 'Density Vs Residual Sugar')
```

The graph displays some interesting features such as the residual sugar is 
increased the density also increases thus making the wine sweeter.Very few have 
residual sugar as  g/dm^3^.However between many wines are made between 
1 and 3 gm/litre.

```{r TSD vs FSD,warning=FALSE,message=FALSE}
ggplot(data=White_wine,aes(x=total.sulfur.dioxide,y=free.sulfur.dioxide))+
  geom_point(alpha=1/50)+
  geom_smooth()+
  coord_cartesian(xlim = c(50,250))+
  xlab('Total sulphur dioxide') +
  ylab('Free Sulphur Dioxide') +
  labs(title = 'Total Sulphur Dioxide vs Free Sulphur Dioxide')
```

Having correlation having 0.62 it has a moderate correlation and it is possible 
since Free sulphuric dioxide and Total sulphuric dioxide are related hence have 
a high correlation.

```{r Density vs TSD,warning=FALSE,message=FALSE}
ggplot(data=White_wine,aes(x=total.sulfur.dioxide,y=density))+
  geom_point(alpha=1/30)+
  geom_smooth()+
  scale_x_log10()+
  scale_x_continuous(limits = c(75,200))+
  coord_cartesian(xlim = c(75,200))+
  xlab('Total Sulphur Dioxide') +
  ylab('Density of Wine') +
  labs(title = 'Density Vs Total Sulphur Dioxide')
```

Points are denser between 80-120 g/dm^3^ and although there is linear increase 
it keeps getting lesser when density is reaching towards 1 g/cm^3^.

##Inference
Alcohol can be considered as a factor that can affect the quality of wine.
Thus creating a graph between Wine Quality and Alcohol is important.

```{r Wine Quality vs Alcohol,warning=FALSE,message=FALSE}
ggplot(aes(category, alcohol), 
       data = White_wine) +
  geom_jitter(alpha=.15)+
  geom_boxplot( alpha = .5,color = 'red')+
  stat_summary(fun.y = "mean", 
              geom = "point", 
             color = "green", 
            shape = 8, 
           size = 4) +
  xlab('Wine Quality') +
  ylab('Alcohol Percentage') +
  labs(title = 'Alcohol Percentage Vs Wine Quality')+
  theme_economist()

```

The Wine Quality was divided into 4 classes i.e. 'Bad','Average','Good',
'Excellent'.A layer of boxplot was placed over point graph.
[link](goo.gl/KBHik6) 
It shows that mean alcohol percentage for Excellent wines is approx 11.8% while
the bad and average wines have approx 10.2 %.  

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in 
the dataset?

Although correlation is not causation but it can help us to determine the 
relationships between  variables easily hence the following variables were taken
and corresponding values of correlation was noted.

    Comparison            | Correlation
    ----------------------|-----------------
    Alcohol/Chlorides     | -0.36
    Alcohol/TSD           | -0.45
    Alcohol/Residual Sugar| -0.45
    Alcohol/Density       | -0.78
    Density/Residual Sugar| 0.84
    TSD/FSD               | 0.62
    Density/TSD           | 0.53
    Alcohol/Quality       | 0.44

### Did you observe any interesting relationships between the other features \
 (not the main feature(s) of interest)?

The relationship between density and TSD shows a correlation of 0.53.There is a
linear relationship between 0.99 to 1.00 as the TSD increases.


### What was the strongest relationship you found?
Strongest relationship was found between Density and Residual Sugar but still 
needs to be investigated upon. 

# Multivariate Plots Section


```{r  Residual Sugar Vs Density(Rating),message=FALSE,warning=FALSE}
ggplot(aes(x = residual.sugar, y = density, color = category),
               data = White_wine) + 
     scale_color_brewer(type = 'seq', 
                        guide = guide_legend(title = 'Rating'),
                        palette = 5) +
     geom_jitter(size = 2,alpha=1/5) +
     geom_smooth(method = "lm", 
                 se = FALSE,
                 size=1)+
     xlim(0,20)+
     ylim(0.985,1.005)+
     xlab('Residual sugar of Wines')+
     ylab('Density of Wines')+
     labs(title='Graph between Residual sugar and 
          Density compared with Category of wines')

```

The previous graphs showed only the linear relationship between Residual Sugar 
and Density of wines.In this graph,the points were separated with respected to 
their rating and thus gives interesting points.These can help us by creating the 
perfect wine.Any residual sugar above 15 is not good for making wine.And density
should be between the ranges of 0.990-0.995.

```{r Density Vs Alcohol(Rating),message=FALSE,warning=FALSE}
ggplot(aes(x = density, y = alcohol, color = category),
       data = White_wine) + 
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(title = 'Rating')) +
  geom_jitter(size = 2,
              alpha=1/5) +
  geom_smooth(method = "lm",
              se = FALSE,
              size=1)+
  xlim(0.985,1.005)+
  ylim(7.5,15)+
  xlab('Density of Wines')+
  ylab('Alcohol Percentage')+
  labs(title='Graph between Alcohol and Density compared with Category of wines')

```

Bad Wines have comparatively low precentage of alcohol as compared to Excellent 
wines as displayed by the four lines.As the density is increased the alcohol 
percentage also decreases thus helping us understand the fact that alcohol 
percentage between 11-13% is quite good for making wine.

```{r TSD vs FSD(Rating),warning=FALSE,message=FALSE}
ggplot(aes(x = total.sulfur.dioxide, y = free.sulfur.dioxide, color = category),
       data = White_wine) + 
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(title = 'Rating'),
                     palette = 3) +
  geom_jitter(size = 2,
              alpha=1/5) +
  geom_smooth(method = "lm",
              se = FALSE,
              size=1)+
  coord_cartesian(xlim = c(0,300),
                  ylim = c(0,100))+
  xlab('Total Sulphur Dioxide')+
  ylab('Free Sulphur Dioxide')+
  labs(title='Graph between TSD and FSD compared with Category of wines')

```

The relationship between TSD(Total Sulphur Dioxide) and FSD 
(Free Sulphur Dioxide) has some interesting features as slope of Bad Wine 
is quite higher than avaerage wines as TSD and FSD increases.

```{r TSD vs Density,warning=FALSE,message=FALSE}
ggplot(aes(x = density, y = total.sulfur.dioxide, color = category),
       data = White_wine) + 
  scale_color_brewer(type = 'qual', 
                     guide = guide_legend(title = 'Rating'),
                     palette = 4) +
  geom_jitter(size = 2,
              alpha=1/5) +
  geom_smooth(method = "lm",
              se = FALSE,
              size=1)+
  coord_cartesian(xlim = c(0.985,1.005),ylim = c(0,250))+
  xlab('Density of wines')+
  ylab('Total Sulphur Dioxide')+
  labs(title='Graph between TSD and density compared with Category of wines')

```

Similar graphs were achieved when they were compared with Bad Wine crosses at 
0.995 density and slope has a tremendous increases.Through this graph,we can 
infer that 100-150 g/dm^3^ can help in preparation of excellent wine ranges.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \ 
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Multivariate analysis gives interesting relationships when we enter the 3^rd^ 
variable.Interesting features on bad wines were found as compared to excellent 
wines due to steep slopes of bad wines as they are increased.



------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One,message=FALSE,warning=FALSE}
ggplot(aes(x = residual.sugar, y = density, color = category),
               data = White_wine) + 
     scale_color_brewer(type = 'seq', 
                        guide = guide_legend(title = 'Rating'),
                        palette = 6) +
     geom_jitter(size = 1,
                 alpha=1/10) +
     geom_smooth(method = "lm",
                 se = FALSE,
                 size=1)+
     xlim(0,20)+
     ylim(0.985,1.005)+
     xlab('Residual sugar of Wines')+
     ylab('Density of Wines')+
     labs(title='Graph between Residual sugar and 
          Density compared with Category of wines')
```

### Description One

The graph between Residual Sugar and Density has similar trend for every 
categoryof rating made.The Density range was between 0.99-0.996 for excellent 
wine.

Secondly,All the four categories are converging towards one area thus reaching 
towards 1 gm/dm^3^.

### Plot Two
```{r echo=FALSE, Plot_Two,warning=FALSE,message=FALSE}
ggplot(aes(x = density, y = alcohol, color = category),
       data = White_wine) + 
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(title = 'Rating'),
                     palette = 3) +
  geom_jitter(size = 1,
              alpha=1/5) +
  geom_smooth(method = "lm", 
              se = FALSE,
              size=1)+
  xlim(0.985,1.005)+
  ylim(7.5,15)+
  xlab('Density of Wines')+
  ylab('Alcohol Percentage')+
  labs(title='Graph between Alcohol and Density compared with Category of wines')+
  theme_dark()
```

### Description Two
The alcohol can be considered as main feature for selection of high quality wine.
It has correlation with most of the factors hence for this case it is the graph 
between Density and Alcohol Percentage.It's the percentage between 10-13.2%.

### Plot Three
```{r echo=FALSE, Plot_Three,message=FALSE,warning=FALSE}
ggplot(aes(category, alcohol), 
       data = White_wine) +
  geom_jitter(alpha=.15)+
  geom_boxplot( alpha = .5,color = 'red')+
  stat_summary(fun.y = "mean", 
              geom = "point", 
             color = "green", 
            shape = 8, 
           size = 4) +
  xlab('Wine Quality') +
  ylab('Alcohol Percentage') +
  labs(title = 'Alcohol Percentage Vs Wine Quality')+
  theme_economist()
```

### Description Three

The graph between Alcohol percentage and Wine Quality shows the final trend 
between the graphs.Most of the Wine quality is between Average and Good wines 
while very less wines are present in bad and excellent ans based on that the 
approximate mean of excellent wine is 11.8%.Secondly,the mean  of bad wine is 
10.2%.Good wines and Excellent wines have alcohol percentage between 10.5% and 
12.4%.

# Reflection

This analysis was done on White wine dataset.White wine is a wine whose colour 
can be straw-yellow, yellow-green, or yellow-gold. It is produced by the 
alcoholic fermentation of the non-coloured pulp of grapes,which may have a skin 
of any colour. White wine has existed for at least 2500 years.

There are 4898 rows and 13 variables in this dataset to workupon hence require 
a mixture of univariate,bivariate and multivariate analysis on it.One of the 
problems faced in this dataset is that for Wine Quality the Average had the 
highest no. of data points hence when plotting with geom_point() function it
was overcrowded with the points.Alcohol showed the highest no. of correlation 
and hence can be considered as a factor for creating a linear model.

However,this dataset only takes in data from one particular area and external 
factors like climate change may also affect the quality of wine and hence next 
time maybe it would be better to have a controlled experiment with a minimum of 
10k data points.Collecting data from various regions may also help with the 
analysis.


